Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-23 ThesisJP/Background/background.tex
Removed vector notation.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging. A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality. For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal opening in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted signal will get shaped into a beam of ultrasound by diffraction. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\Delta = \frac{\lambda}{D},
\end{align}
\nomi{$\Delta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength and \nom{$D$}{Aperture width} is the aperture width. With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm. The axial resolution is inverse proportional to the pulse length, hence the frequency bandwidth of the probe.

If time delays are applied on the signal sent to the different elements it is possible to control the steering and focusing of the beam. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied on both transmit and receive.

The standard form of time domain array beamforming is the delay-and-sum beamformer where the output $z$ is calculated as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function. The weight vector is typically selected in order to trade resolution for a lower side lobe level. Hence, the weights are typically real and their K-space response is then symmetric.

>>>>

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging. A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality. For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\delta\theta = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width. With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}. A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in sector which are $\theta$ wide. This kind of image is known as a B-mode image (B for brightness). If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the beamforming technique known as \textit{delay-and-sum} is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is an A-mode image.  If the phase center of our array is the origin and $p$ is the point where we want to steer and focus the ultrasound beam we have to use delays given as:
\begin{align}
\Delta_m[n] =\frac{|x_m+p| - |p|}{c},
\end{align}
where $c$ is the speed of sound.

%The weight vector is typically selected in order to trade resolution for a lower side lobe level. Hence, the weights are typically real and their K-space response is then symmetric.
